Clustering analysis and statistics

### 2019-03-27 notebooks/Clustering analysis.Rmd
minor changes

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtree)
library(cluster)
library(Biostrings)
library(ape)
library(tibble)
library(cowplot)
```

# cgMLST Schema and quality control
```{r import data}
# Vector of excluded samples
ex_samples <- c(
  "2016-17-565-1-S",
  "2016-02-428-2-S",
  "2016-02-486-2-S",
  "2016-02-732-2-S",
  "2014-01-1741-1-S"
  )

@@

@@

```{r Number of missing loci}
boxplot(cgmlst_stats$number.of.missing.data,
        ylab = "Number of missing loci")
```

# Total dendrogram of all isolates
```{r, fig.height=20, fig.width=20, message=FALSE, warning=FALSE}
cgMLST_data <- read.table("../data/chewbbaca/complete_results/clean_cgMLST_data.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

@@

tree_heatmap_data <- read.table("../data/chewbbaca/complete_results/tree_heatmap_data.txt",
                            sep = "\t",
                            header = TRUE)

@@

```{r, fig.height=25, fig.width=25, message=FALSE, warning=FALSE}

@@

palette3 <- c("0" = "grey95",
              # gyrA
              "S83L" = "#c6dbef",
              "S83A" = "#9ecae1",
              "D87N" = "#6baed6",
              "D87H" = "#4292c6",
              "D87Y" = "#2171b5",
              "D87G" = "#08519c",
              "S83L, D87N" = "#08306b",
              # parC
              "S80I" = "#a1d99b",
              "S80R" = "#74c476",
              "S57T" = "#41ab5d",
              "A56T, S80I" = "#238b45",
              "S80I, E84V" = "#005a32",
              # parE
              "D475E" = "#fdd0a2",
              "D463N" = "#fdae6b",
              "S458A" = "#fd8d3c",
              "L416F" = "#f16913",
              "H516Y" = "#d94801",
              "L488M, A512T" = "#8c2d04",
              # PMQR
              "qnrA1" = "#762a83",
              "qnrB19" = "#ef8a62",
              "qnrS1" = "#a6dba0",
              "qnrS2" = "#5aae61",
              "qnrS4" = "#1b7837",
              "qepA4" = "#80cdc1",
              # num
              "1" = "#440154FF",
              "2" = "#471164FF",
              "3" = "#481F70FF",
              "4" = "#472D7BFF",
              "5" = "#443A83FF",
              "6" = "#404688FF",
              "7" = "#3B528BFF",
              "8" = "#365D8DFF",
              "9" = "#31688EFF",
              "10" = "#2C728EFF",
              "11" = "#287C8EFF",
              "12" = "#24868EFF",
              "13" = "#21908CFF",
              "14" = "#1F9A8AFF",
              "15" = "#20A486FF",
              "16" = "#27AD81FF",
              "17" = "#35B779FF",
              "18" = "#47C16EFF",
              "19" = "#5DC863FF",
              "20" = "#75D054FF",
              "21" = "#8FD744FF",
              "22" = "#AADC32FF",
              "23" = "#C7E020FF",
              "24" = "#E3E418FF",
              "25" = "#FDE725FF")

@@

broiler_complete <- gheatmap(broiler_tree_opened,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames_offset_y = 0.75,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

@@

bird_tree_annotated <- ggtree(bird_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#f46d43",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.01,
               size = 5.5)

@@

# Total plot
tot_plot <- plot_grid(broiler_complete,
          pig_complete,
          fox_complete,
          bird_complete,
          nrow = 2,
          ncol = 2,
          align = "h")
tot_plot

@@

# Statistics
```{r, warning=FALSE, message=FALSE}
alleles <- cgMLST_allele_data %>%
  left_join(isolate_data[,c("id","species")], by = "id") %>%
  group_by(species) %>%
  mutate(n = 1:n(),
         new_id = case_when(species == "Broiler" ~ paste("BR", n, sep = "_"),
                            species == "Pig" ~ paste("PI", n, sep = "_"),
                            species == "Red fox" ~ paste("RF", n, sep = "_"),
                            species == "Wild bird" ~ paste("WB", n, sep = "_"))) %>%
  ungroup() %>%
  select(new_id, everything(), -c(species, id)) %>%
  mutate_all(funs(as.factor)) %>%
  column_to_rownames("new_id")

run_df <- lower_tri_dist_calc(alleles, data_frame = TRUE) %>%
  rownames_to_column("id") %>%
  gather(isolate1, value, -id) %>%
  dplyr::rename("isolate2" = id) %>%
  mutate(group1 = substr(isolate1, start = 1, stop = 2),
         group2 = substr(isolate2, start = 1, stop = 2)) %>%
  select(isolate1, group1, isolate2, group2, value) %>%
  filter(is.na(value) == FALSE,
         group1 == group2) %>%
  arrange(group1, group2)

below_calc <- function(df, run, dist_value = 0.1) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = value < dist_value) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Below", "Above")) %>%
    spread(n, nn) %>%
    mutate(Below = ifelse(is.na(Below) == TRUE, 0, Below)) %>%
    mutate(Percent_below = round(Below / (Above + Below) * 100, 2),
           iter = run)
  return(df)
}

below_calc2 <- function(df, run, dist_value = 0.1) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = n()) %>%
    sample_n(min(.$n)) %>%
    select(-n) %>%
    mutate(n = value < dist_value) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Below", "Above")) %>%
    spread(n, nn) %>%
    mutate(Below = ifelse(is.na(Below) == TRUE, 0, Below)) %>%
    mutate(Percent_below = round(Below / (Above + Below) * 100, 2),
           iter = run)
  return(df)
}

above_calc <- function(df, run, dist_value = 0.9) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = value > dist_value) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Above", "Below")) %>%
    spread(n, nn) %>%
    mutate(Above = ifelse(is.na(Above) == TRUE, 0, Above)) %>%
    mutate(Percent_above = round(Above / (Above + Below) * 100, 2),
           iter = run)
  return(df)
}

above_calc2 <- function(df, run, dist_value = 0.9) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = n()) %>%
    sample_n(min(.$n)) %>%
    select(-n) %>%
    mutate(n = value > dist_value) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Above", "Below")) %>%
    spread(n, nn) %>%
    mutate(Above = ifelse(is.na(Above) == TRUE, 0, Above)) %>%
    mutate(Percent_above = round(Above / (Above + Below) * 100, 2),
           iter = run)
  return(df)
}

mean_calc <- function(df, run) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(metric = round(mean(value), 2)) %>%
    select(group1, group2, metric) %>%
    summarise_all(funs(func_paste)) %>%
    mutate(iter = run,
           metric = as.numeric(metric))
}

median_calc <- function(df, run) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(metric = round(median(value), 2)) %>%
    select(group1, group2, metric) %>%
    summarise_all(funs(func_paste)) %>%
    mutate(iter = run,
           metric = as.numeric(metric))
}

quantile_calc <- function(df, run) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(quant_0 = quantile(value)[1],
           quant_25 = quantile(value)[2],
           quant_50 = quantile(value)[3],
           quant_75 = quantile(value)[4],
           quant_100 = quantile(value)[5]) %>%
    select(group1, group2, contains("quant")) %>%
    summarise_all(funs(func_paste)) %>%
    mutate_at(vars(contains("quant")),
              funs(as.numeric)) %>%
    mutate(iter = run)
}

iterate_samples <- function(df, run, method = "mean", dist_value) {

  df <- df %>%
    mutate(value = sample(value))

    if (method == "mean") {
      df <- mean_calc(df, run)
    }

    if (method == "below") {
     df <- below_calc2(df, run, dist_value)
    }

    if (method == "above") {
     df <- above_calc2(df, run, dist_value)
    }

@@

# run iterations on selected data frame
run_iterations <- function(df, runs = 1000, seed = 10, method = "mean", dist_value) {
  # set seed
  set.seed(seed)

@@

    if (method == "below") {
     orig <- below_calc(df, 0, dist_value)
    }

   if (method == "above") {
     orig <- above_calc(df, 0, dist_value)
    }

@@

  # run iterations
  for (i in 1:runs) {
    result <- iterate_samples(df,
                              i,
                              method = method,
                              dist_value = dist_value)
    output <- c(output, list(result))
  }
  return(output)
}

# create distribution plots from iteration data
create_plot <- function(df, group_one, column, exp_value, title, binwidth = 0.2) {

  df %>%
    filter(group1 == group_one) %>%
    ggplot(aes(!! sym(column))) +
    geom_histogram(binwidth = 0.2, color = "black",
                   fill = "grey80") +
    geom_segment(aes(x = exp_value, xend = exp_value, y = 70, yend = 0),
                 arrow = arrow(length = unit(0.5, "cm"))) +
    annotate("text", x = exp_value, y = 75, label = "Expected value", angle = 90, size = 3, hjust = 0) +
    ggtitle(title)
}

# Run functions
complete_data_below <- run_iterations(run_df,
                                runs = 1000,
                                method = "below",
                                dist_value = 0.1) %>%
  bind_rows()

complete_data_above <- run_iterations(run_df,
                                runs = 1000,
                                method = "above",
                                dist_value = 0.8) %>%
  bind_rows()

@@

p1 <- ggplot(complete_data_below, aes(Percent_below)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, color = "black",
                fill = "grey80") +
  stat_function(fun = dnorm,
                args = with(complete_data_below, c(mean = mean(Percent_below), sd = sd(Percent_below))),
                color = "red") +
  geom_segment(aes(x = complete_data_below$Percent_below[1],
                   xend = complete_data_below$Percent_below[1], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = complete_data_below$Percent_below[2],
                   xend = complete_data_below$Percent_below[2], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = complete_data_below$Percent_below[3],
                   xend = complete_data_below$Percent_below[3], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = complete_data_below$Percent_below[4],
                   xend = complete_data_below$Percent_below[4], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#fdae61") +
  geom_vline(xintercept = mean(complete_data_below$Percent_below),
             size = 1)

p2 <- ggplot(complete_data_above, aes(Percent_above)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, color = "black",
                fill = "grey80") +
  stat_function(fun = dnorm,
                args = with(complete_data_above, c(mean = mean(Percent_above), sd = sd(Percent_above))),
                color = "red") +
  geom_segment(aes(x = complete_data_above$Percent_above[1],
                   xend = complete_data_above$Percent_above[1], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = complete_data_above$Percent_above[2],
                   xend = complete_data_above$Percent_above[2], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = complete_data_above$Percent_above[3],
                   xend = complete_data_above$Percent_above[3], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = complete_data_above$Percent_above[4],
                   xend = complete_data_above$Percent_above[4], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#fdae61") +
  geom_vline(xintercept = mean(complete_data_above$Percent_above),
             size = 1)

plot_grid(p1, p2, ncol = 2, nrow = 1, align = "v")

mean_data_below <- complete_data_below %>%
  mutate(mean = mean(Percent_below),
         dist_from_mean = mean - Percent_below)

mp1 <- ggplot(mean_data_below, aes(dist_from_mean))+
  geom_histogram(binwidth = 0.15,
                 color = "black") +
  geom_segment(aes(x = mean_data_below$dist_from_mean[1],
                   xend = mean_data_below$dist_from_mean[1], y = 100, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = mean_data_below$dist_from_mean[2],
                   xend = mean_data_below$dist_from_mean[2], y = 100, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = mean_data_below$dist_from_mean[3],
                   xend = mean_data_below$dist_from_mean[3], y = 100, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = mean_data_below$dist_from_mean[4],
                   xend = mean_data_below$dist_from_mean[4], y = 100, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#fdae61")

mean_data_above <- complete_data_above %>%
  mutate(mean = mean(Percent_above),
         dist_from_mean = mean - Percent_above)

mp2 <- ggplot(mean_data_above, aes(dist_from_mean))+
  geom_histogram(binwidth = 0.15,
                 color = "black") +
  geom_segment(aes(x = mean_data_above$dist_from_mean[1],
                   xend = mean_data_above$dist_from_mean[1], y = 50, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = mean_data_above$dist_from_mean[2],
                   xend = mean_data_above$dist_from_mean[2], y = 50, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = mean_data_above$dist_from_mean[3],
                   xend = mean_data_above$dist_from_mean[3], y = 50, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = mean_data_above$dist_from_mean[4],
                   xend = mean_data_above$dist_from_mean[4], y = 50, yend = 0),
               arrow = arrow(length = unit(0.4, "cm"),
                             type = "closed"),
               color = "#fdae61")

plot_grid(mp1, mp2, ncol = 2, nrow = 1, align = "v")

chsq_data_below <- complete_data_below[1:4,] %>%
  mutate(Total = Above + Below)

br_pi_csq <- chsq_data_below[1:2,] %>%
  select(Total, Below) %>%
  as.matrix

br_rf_csq <- chsq_data_below[c(1,3),] %>%
  select(Total, Below) %>%
  as.matrix

br_wb_csq <- chsq_data_below[c(1,4),] %>%
  select(Total, Below) %>%
  as.matrix

pi_rf_csq <- chsq_data_below[c(2,3),] %>%
  select(Total, Below) %>%
  as.matrix

pi_wb_csq <- chsq_data_below[c(2,4),] %>%
  select(Total, Below) %>%
  as.matrix

rf_wb_csq <- chsq_data_below[c(3,4),] %>%
  select(Total, Below) %>%
  as.matrix

chisq.test(br_pi_csq, correct = FALSE)
chisq.test(br_rf_csq, correct = FALSE)
chisq.test(br_wb_csq, correct = FALSE)
chisq.test(pi_rf_csq, correct = FALSE)
chisq.test(pi_wb_csq, correct = FALSE)
chisq.test(rf_wb_csq, correct = FALSE)

@@

>>>>

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtree)
library(cluster)
library(Biostrings)
library(ape)
library(tibble)
library(cowplot)
library(knitr)
library(png)
```

# Core genome MLST schema and quality control
The _E. coli_ scheme from Enterobase [1] was used with chewBBACA [2] to get the cgMLST profile for each isolate.
The number of missing loci in total (sum of LNF, NIPH, PLOT, ALM and ASM from chewBBACA) is plotted below for all isolates.

```{r import data}
# Vector of excluded samples
ex_samples <- c(
  "2016-17-565-1-S",
  "2016-02-428-2-S",
  "2016-02-486-2-S",
  "2016-02-732-2-S",
  "2014-01-1741-1-S"
  )

@@

detailed_cgmlst_stats <- read.table("../data/chewbbaca/results_statistics.tsv",
                                    sep = "\t",
                                    header = TRUE,
                                    stringsAsFactors = FALSE)

@@

```{r cgMLST stats}
boxplot(cgmlst_stats$number.of.missing.data,
        ylab = "Number of missing loci")
```

Below follows the detailed results for all loci in the cgMLST scheme. EXC = excact matches, INF = novel allele, LNF = locus not found, NIPH = paralogous locus, PLOT = possible loci on tip of contig, ALM = allele 20 % larger than reference, ASM = allele 20 % smaller than reference. The numbers below each boxplot denote the mean.

```{r}
test <- detailed_cgmlst_stats %>%
  gather(metric, value, -Genome) %>%
  group_by(metric) %>%
  mutate(mean_val = round(mean(value), 1))

ggplot(test, aes(metric, value)) +
  stat_boxplot(geom = "errorbar", width = 0.5) +
  geom_boxplot() +
  geom_text(data = subset(test,
                          Genome == "2006-01-2184_S30_pilon_spades.fasta"),
            aes(label = mean_val),
            y = -60,
            size = 3.8) +
  scale_x_discrete(limits = c("EXC",
                              "INF",
                              "LNF",
                              "NIPH",
                              "PLOT",
                              "ALM",
                              "ASM")) +
  labs(y = "Number of loci") +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank())
```

# Total dendrogram of all isolates
Genetic distances were calculated from the allele data from the cgMLST report, and a dendrogram was drawn from these distances. The annotations are based on data from the ARIBA resistance gene analysis.
```{r, fig.height=20, fig.width=20, message=FALSE, warning=FALSE}
cgMLST_data <- read.table("../data/chewbbaca/complete_results/clean_cgMLST_data.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

@@

tree_heatmap_data <- read.table("../data/chewbbaca/complete_results/tree_heatmap_data.txt",
                            sep = "\t",
                            header = TRUE) %>%
  select(-gadX)

@@

# Dendrograms per animal species
```{r, fig.height=25, fig.width=25, message=FALSE, warning=FALSE}

@@

palette3 <- c("0" = "grey95",
              # gyrA
              "S83L" = "#c6dbef",
              "S83A" = "#9ecae1",
              "D87N" = "#6baed6",
              "D87H" = "#4292c6",
              "D87Y" = "#2171b5",
              "D87G" = "#08519c",
              "S83L, D87N" = "#08306b",
              # parC
              "S80I" = "#a1d99b",
              "S80R" = "#74c476",
              "S57T" = "#41ab5d",
              "A56T, S80I" = "#238b45",
              "S80I, E84V" = "#005a32",
              # parE
              "D475E" = "#fdd0a2",
              "D463N" = "#fdae6b",
              "S458A" = "#fd8d3c",
              "L416F" = "#f16913",
              "H516Y" = "#d94801",
              "L488M, A512T" = "#8c2d04",
              # PMQR
              "qnrA1" = "#762a83",
              "qnrB19" = "#ef8a62",
              "qnrS1" = "#a6dba0",
              "qnrS2" = "#5aae61",
              "qnrS4" = "#1b7837",
              "qepA4" = "#80cdc1",
              # num
              "1" = "#440154FF",
              "2" = "#3B528BFF",
              "3" = "#21908CFF",
              "4" = "#5DC863FF",
              "5" = "#FDE725FF")

@@

broiler_complete <- gheatmap(broiler_tree_opened,
         metadata_species,
         offset = 0.1,
         width = 0.5,
         colnames_offset_y = 0.75,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

@@

bird_tree_annotated <- ggtree(bird_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#fdae61",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.01,
               size = 5.5)

@@

# Total plot
tot_plot <- plot_grid(broiler_complete,
          pig_complete,
          fox_complete,
          bird_complete,
          nrow = 2,
          ncol = 2,
          align = "h")

include_graphics("../figures/species_dendrogram.png")
```

# Statistics
## Distribution of calculated distances per animal species
Below follows a plot of the distribution of the lower triangular distances for each animal species. A bimodal distribution is present in all four plots. The filtering values of < 0.1 and > 0.8 is used for downstream statistics to acertain whether the broiler isolates are more homogenous than the isolates from other species.

@@

## Distribution of iterated distances
Distribution of percentages of distances < 0.1 or > 0.8. The distribution is based on 1000 iterations of the distance data. Observed values are denoted as arrows for each animal species; dark blue = Broiler, light blue = Pig, red = Red fox, yellow = Wild bird. The mean value is denoted as a vertical black line.
```{r, warning=FALSE, message=FALSE, dpi=300}
# generate data frame from allele data
alleles <- cgMLST_allele_data %>%
  left_join(isolate_data[,c("id","species")], by = "id") %>%
  group_by(species) %>%
  mutate(n = 1:n(),
         new_id = case_when(species == "Broiler" ~ paste("BR", n, sep = "_"),
                            species == "Pig" ~ paste("PI", n, sep = "_"),
                            species == "Red fox" ~ paste("RF", n, sep = "_"),
                            species == "Wild bird" ~ paste("WB", n, sep = "_"))) %>%
  ungroup() %>%
  select(new_id, everything(), -c(species, id)) %>%
  mutate_all(funs(as.factor)) %>%
  column_to_rownames("new_id")

# generate data frame for iteration analysis
run_df <- lower_tri_dist_calc(alleles, data_frame = TRUE) %>%
  rownames_to_column("id") %>%
  gather(isolate1, value, -id) %>%
  dplyr::rename("isolate2" = id) %>%
  mutate(group1 = substr(isolate1, start = 1, stop = 2),
         group2 = substr(isolate2, start = 1, stop = 2)) %>%
  select(isolate1, group1, isolate2, group2, value) %>%
  filter(is.na(value) == FALSE,
         group1 == group2) %>%
  arrange(group1, group2)

# functions used in iteration analyses

## Calculates the sum of observations lower than dist_value1 and
## bigger than dist_value2, and calculates the percentage of
## observed values for each group
without <- function(df, run, dist_value1 = 0.1, dist_value2 = 0.8) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = value < dist_value1 | value > dist_value2) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Within", "Without")) %>%
    spread(n, nn) %>%
    mutate(Percent = round(Within / (Within + Without) * 100, 2),
           iter = run)
  return(df)
}

## Calculates the sum of observations between dist_value1 and dist_value2
## and calculates the percentage of observed values for each group
within <- function(df, run, dist_value1 = 0.1, dist_value2 = 0.8) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = value > dist_value1 & value < dist_value2) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Within", "Without")) %>%
    spread(n, nn) %>%
    mutate(Percent = round(Within / (Within + Without) * 100, 2),
           iter = run)
  return(df)
}

## Calculates the sum of observations below dist_value1
## for each group
below_calc <- function(df, run, dist_value1 = 0.1) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = value < dist_value1) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Below", "Above")) %>%
    spread(n, nn) %>%
    mutate(Below = ifelse(is.na(Below) == TRUE, 0, Below)) %>%
    mutate(Percent_below = round(Below / (Above + Below) * 100, 2),
           iter = run)
  return(df)
}

## Calculates the sum of observations above dist_value1
## for each group
above_calc <- function(df, run, dist_value1 = 0.9) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(n = value > dist_value1) %>%
    group_by(group1, group2, n) %>%
    count() %>%
    ungroup() %>%
    mutate(n = ifelse(n == TRUE, "Above", "Below")) %>%
    spread(n, nn) %>%
    mutate(Above = ifelse(is.na(Above) == TRUE, 0, Above)) %>%
    mutate(Percent_above = round(Above / (Above + Below) * 100, 2),
           iter = run)
  return(df)
}

## Calculates the mean of observations for each group
mean_calc <- function(df, run) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(metric = round(mean(value), 2)) %>%
    select(group1, group2, metric) %>%
    summarise_all(funs(func_paste)) %>%
    mutate(iter = run,
           metric = as.numeric(metric))
}

## Calculates the median of observations for each group
median_calc <- function(df, run) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(metric = round(median(value), 2)) %>%
    select(group1, group2, metric) %>%
    summarise_all(funs(func_paste)) %>%
    mutate(iter = run,
           metric = as.numeric(metric))
}

## Calculates the quantiles of the observations for each group
quantile_calc <- function(df, run) {
  df <- df %>%
    group_by(group1, group2) %>%
    mutate(quant_0 = quantile(value)[1],
           quant_25 = quantile(value)[2],
           quant_50 = quantile(value)[3],
           quant_75 = quantile(value)[4],
           quant_100 = quantile(value)[5]) %>%
    select(group1, group2, contains("quant")) %>%
    summarise_all(funs(func_paste)) %>%
    mutate_at(vars(contains("quant")),
              funs(as.numeric)) %>%
    mutate(iter = run)
}

# Iteration function
iterate_samples <- function(df, run, method = "mean", dist_value1, dist_value2) {

  # randomize the value column, regardless of groups
  df <- df %>%
    mutate(value = sample(value))

  # run specified function
  if (method == "mean") {
    df <- mean_calc(df, run)
    }

  if (method == "below") {
   df <- below_calc(df, run, dist_value1)
    }

  if (method == "above") {
   df <- above_calc(df, run, dist_value1)
    }

  if (method == "within") {
    df <- within(df, run, dist_value1, dist_value2)
    }

  if (method == "without") {
    df <- without(df, run, dist_value1, dist_value2)
    }

@@

# run iterations on selected data frame
run_iterations <- function(df, runs = 1000, seed = 10, method = "mean", dist_value1, dist_value2) {
  # set seed
  set.seed(seed)

@@

  if (method == "below") {
   orig <- below_calc(df, 0, dist_value1)
    }

  if (method == "above") {
   orig <- above_calc(df, 0, dist_value1)
    }

  if (method == "within") {
    orig <- within(df, 0, dist_value1, dist_value2)
    }

  if (method == "without") {
    orig <- without(df, 0, dist_value1, dist_value2)
    }

@@

  # run iterations
  for (i in 1:runs) {
    result <- iterate_samples(df,
                              i,
                              method = method,
                              dist_value1 = dist_value1,
                              dist_value2 = dist_value2)
    output <- c(output, list(result))
    }
  return(output)
}

# Run functions
summary_stats <- run_df %>%
  mutate(total_mean = round(mean(value)*100, 3)) %>%
  group_by(group1) %>%
  mutate(mean = round(mean(value)*100, 3),
         var = round(var(value)*100, 3)) %>%
  select(group1, total_mean, mean, var) %>%
  summarise_all(funs(func_paste))

summary_stats2 <- run_df %>%
  mutate(total_mean = round(mean(value), 3)) %>%
  group_by(group1) %>%
  mutate(mean = round(mean(value), 3),
         var = round(var(value), 3)) %>%
  select(group1, total_mean, mean, var) %>%
  summarise_all(funs(func_paste))

complete_data <- run_iterations(run_df,
                                runs = 1000,
                                method = "without",
                                dist_value1 = 0.1,
                                dist_value2 = 0.8) %>%
  bind_rows() %>%
  mutate(Percent = as.numeric(Percent))

lower_data <- run_iterations(run_df,
                             runs = 1000,
                             method = "below",
                             dist_value1 = 0.1) %>%
  bind_rows() %>%
  mutate(Percent_below = as.numeric(Percent_below))

@@

ggplot(complete_data, aes(Percent)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, color = "black",
                fill = "grey80") +
  stat_function(fun = dnorm,
                args = with(complete_data, c(mean = mean(Percent), sd = sd(Percent))),
                color = "red") +
  geom_segment(aes(x = complete_data$Percent[1],
                   xend = complete_data$Percent[1], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = complete_data$Percent[2],
                   xend = complete_data$Percent[2], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = complete_data$Percent[3],
                   xend = complete_data$Percent[3], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = complete_data$Percent[4],
                   xend = complete_data$Percent[4], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#fdae61") +
  geom_text(label = "p < 0.001",
            x = 83.30,
            y = 0.16,
            angle = 90) +
  geom_vline(xintercept = as.numeric(summary_stats$mean[1]),
             size = 1,
             color = "#4575b4") +
  geom_vline(xintercept = as.numeric(summary_stats$mean[2]),
             size = 1,
             color = "#74add1") +
  geom_vline(xintercept = as.numeric(summary_stats$mean[3]),
             size = 1,
             color = "#f46d43") +
  geom_vline(xintercept = as.numeric(summary_stats$mean[4]),
             size = 1,
             color = "#fdae61") +
  labs(x = "Percent (%) distances < 0.1 or > 0.8",
       y = "Density") +
  ggtitle(label = "Distribution of percentages",
          subtitle = "Number of iterations: 1000") +
  theme(plot.title = element_text(hjust = 0))

ggplot(lower_data, aes(Percent_below)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, color = "black",
                fill = "grey80") +
  stat_function(fun = dnorm,
                args = with(lower_data, c(mean = mean(Percent_below), sd = sd(Percent_below))),
                color = "red") +
  geom_segment(aes(x = lower_data$Percent_below[1],
                   xend = lower_data$Percent_below[1], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = lower_data$Percent_below[2],
                   xend = lower_data$Percent_below[2], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = lower_data$Percent_below[3],
                   xend = lower_data$Percent_below[3], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = lower_data$Percent_below[4],
                   xend = lower_data$Percent_below[4], y = 0.15, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#fdae61") +
  geom_text(label = "p < 0.001",
            x = 5.87,
            y = 0.32,
            angle = 90) +
  geom_vline(xintercept = mean(lower_data$Percent_below),
             size = 1) +
  labs(x = "Percent (%) distances < 0.1",
       y = "Density") +
  ggtitle(label = "Distribution of Percentages",
          subtitle = "Number of iterations: 1000") +
  theme(plot.title = element_text(hjust = 0))

summary_stats2

@@

# References
[1] Alikhan N-F, Zhou Z, Sergeant MJ, Achtman M. A genomic overview of the population structure of Salmonella. Casadesús J, editor. PLOS Genet [Internet]. 2018 Apr 5;14(4):e1007261. Available from: https://dx.plos.org/10.1371/journal.pgen.1007261

[2] Machado MP, Silva M, Carriço JA, Silva DN, Santos S, Ramirez M, et al. chewBBACA: A complete suite for gene-by-gene schema creation and strain identification. Microb Genomics. 2018;1–7.
